---
output:
  github_document:
    toc: false
    html_preview: false
---

<!-- Generated as README.md from README.Rmd. Edit README.Rmd. -->

# Rdragmap

<!-- badges: start -->
[![R-CMD-check](https://github.com/RGenomicsETL/Rdragmap/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/RGenomicsETL/Rdragmap/actions/workflows/R-CMD-check.yaml)
[![R-universe](https://rgenomicsetl.r-universe.dev/badges/Rdragmap)](https://rgenomicsetl.r-universe.dev/Rdragmap)
<!-- badges: end -->

Rdragmap builds the portable Rdragmap `dragen-os` executable during package
installation and invokes that installed executable through explicit argument
vectors. It never searches `PATH` and does not expose DRAGMAP's internal static
archives as an R ABI.

The wrapper currently supports DRAGMAP v8 reference generation and FASTQ to
SAM alignment. Native behavior and intentional native corrections are documented
in the parent repository's `CONFORMANCE.md` and `ERRATA.md`.

## Installation requirements

Installation requires a Unix-like system, GNU Make, a C++17 compiler, zlib,
bzip2, and development files for Boost Iostreams and Program Options. The
package's `configure` script uses the compiler and make program reported by R,
then builds only the package-owned `dragen-os` child executable.

## Use

All file paths are absolute. Build an index into a directory that does not yet
exist:

```{r build-index, eval=FALSE}
library(Rdragmap)

native <- rdragmap_executables()
built <- rdragmap_build_index(
  reference_fasta = "/data/reference.fa",
  index_directory = "/data/reference.rdragmap",
  executables = native,
  threads = 2L
)
stopifnot(!rdragmap_is_error(built))
```

Align single-end or paired FASTQ. SAM, native mapping metrics, and, for paired
input, insert-size statistics are written beside `output_sam`.

```{r align, eval=FALSE}
aligned <- rdragmap_align(
  index = built@index,
  read1 = "/data/reads_1.fastq.gz",
  read2 = "/data/reads_2.fastq.gz",
  output_sam = "/data/sample.sam",
  executables = native,
  read_group_id = "sample-1",
  sample_name = "sample-1",
  threads = 2L
)
stopifnot(!rdragmap_is_error(aligned))
```

## Installed miniature example

The package ships a tiny FASTA and FASTQ for a complete local smoke workflow.
It creates all generated files under a temporary directory and removes them
when finished.

```{r miniature-example, eval=FALSE}
work <- tempfile("rdragmap-example-")
dir.create(work)

built <- rdragmap_build_index(
  reference_fasta = system.file("extdata", "tiny.fasta", package = "Rdragmap"),
  index_directory = file.path(work, "index"),
  threads = 1L,
  hash_size = "16MB"
)
stopifnot(!rdragmap_is_error(built))

aligned <- rdragmap_align(
  index = built@index,
  read1 = system.file("extdata", "one.fastq", package = "Rdragmap"),
  output_sam = file.path(work, "tiny.sam"),
  read_group_id = "tiny",
  sample_name = "tiny",
  threads = 1L,
  enable_sampling = FALSE
)
stopifnot(!rdragmap_is_error(aligned))
unlink(work, recursive = TRUE, force = TRUE)
```

Expected environmental and process failures are returned as
`RdragmapErrorValue` subclasses with stable `@code` fields. Invalid argument
contracts signal `rdragmap_contract_violation()` conditions.

## Development

From this package directory:

```sh
make bootstrap
make rd
make build
make test
make check
```

`make bootstrap` creates `tools/dragmap-source.tar.xz`, the exact native source
closure used by `configure`. `make source-audit` verifies its members against
the source manifest.
