---
output:
  github_document:
    toc: false
    html_preview: false
---

<!-- Generated as README.md from README.Rmd. Edit README.Rmd and the benchmark CSV files. -->

# hs37d5 and HG02088 exome validation

This is an executed compatibility and performance check, not a bundled data
fixture. Large inputs and generated indexes remain outside Git.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
timings <- read.csv(
  "performance.csv",
  stringsAsFactors = FALSE,
  check.names = FALSE
)
profiles <- read.csv(
  "profile.csv",
  stringsAsFactors = FALSE,
  check.names = FALSE
)
threading <- read.csv(
  "threading.csv",
  stringsAsFactors = FALSE,
  check.names = FALSE
)
format_timing_table <- function(rows, label_name) {
  labels <- rows$implementation
  data.frame(
    setNames(list(labels), label_name),
    `Wall time` = sprintf("%.2f s", rows$wall_seconds),
    `User CPU` = sprintf("%.2f s", rows$user_cpu_seconds),
    `System CPU` = sprintf("%.2f s", rows$system_cpu_seconds),
    `CPU utilization` = ifelse(
      is.na(rows$cpu_utilization_percent),
      "",
      sprintf("%d%%", rows$cpu_utilization_percent)
    ),
    `Peak RSS` = paste0(
      format(rows$peak_rss_kib, big.mark = ",", scientific = FALSE, trim = TRUE),
      " KiB"
    ),
    check.names = FALSE
  )
}
```

## Authorities

Validation was run on 2026-08-05 with:

- untouched DRAGMAP commit
  `4f98e00e2aedc85e27ea6c118cf7b16663036c14`;
- Rdragmap commit `b7c5756561fcf5fe335d8e7a37dd5b3e3722c958`;
- 16 mapper or hash-builder threads;
- Ubuntu 24.04, Linux 6.8, 62 GiB RAM, and an Intel Core i5-13500
  (20 logical CPUs, AVX2 available).

The reference was the 1000 Genomes hs37d5 FASTA:

- URL: <https://ftp.1000genomes.ebi.ac.uk/vol1/ftp/technical/reference/phase2_reference_assembly_sequence/hs37d5.fa.gz>
- compressed SHA-256:
  `e9157e19a95e01dfc47080b5b6aa559c861de90b9934c2ea7c49cd5ec49e0285`
- 86 sequences and 3,137,454,505 unpadded bases.

The read authority was the real 1000 Genomes phase 3 exome run SRR716421
from HG02088. The project sequence index classifies both mates as `exome`:
<https://ftp.1000genomes.ebi.ac.uk/vol1/ftp/phase3/20130502.phase3.sequence.index>.

- mate 1 URL: <https://ftp.1000genomes.ebi.ac.uk/vol1/ftp/phase3/data/HG02088/sequence_read/SRR716421_1.filt.fastq.gz>
- mate 2 URL: <https://ftp.1000genomes.ebi.ac.uk/vol1/ftp/phase3/data/HG02088/sequence_read/SRR716421_2.filt.fastq.gz>
- declared MD5: `bc97e5ddcd90f36bbe98324718b66a27` and
  `54e095386d935fc95e7ce25d32b3110d` (both verified);
- observed SHA-256: `baa696169f804ae46c468e9c69bb058fc41a753c17fdcdc23168ca8a3f186232`
  and `090ac1f91c97fb73441624aee8b5863afde1648c51939b3af608c11ef6321823`;
- 10,110,535 records in each mate, with gzip integrity verified.

## Hash-table compatibility

Both binaries built hs37d5 at the same canonical output path with:

```sh
dragen-os \
  --build-hash-table true \
  --ht-reference hs37d5.fa \
  --output-directory INDEX \
  --ht-num-threads 16 \
  --ht-write-hash-bin 0
```

The following generated files were byte-identical:

- `hash_table.cmp` (`c835d1cf...10e45`);
- `reference.bin` (`0c23c0f8...37701`);
- `ref_index.bin` (`df020e03...edffa`);
- `repeat_mask.bin` (`1383eb3c...f12ba`);
- `str_table.bin` (`c077f618...8576`).

All text configuration fields after the generated command line were identical,
including:

```text
digest               = 0xDA20A269
hash_digest          = 0xC7BA64EA
extend_table_digest  = 0x12663D7F
```

`hash_table.cfg.bin` differed only in its embedded generating command string,
because the executable paths have different lengths. `hash_table_stats.txt`
matched before its intentionally volatile cycle-counter section. Untouched
DRAGMAP successfully decompressed and mapped against the Rdragmap-produced
index.

A first full-reference run exposed that conventional complemented CRC-32C is
not the raw state transition used by the imported `crc32q` hash digest. Commit
`a5e8e9d` corrected that distinction; `b7c5756` uses the exact fixed-size
portable transition. A million chained random 64-bit transitions and fixed
regression vectors match the x86 `crc32q` oracle.

## Measured hash-build performance

```{r hash-build-table, results='asis'}
hash_rows <- subset(timings, phase == "hash-build")
knitr::kable(format_timing_table(hash_rows, "Build"), align = c("l", rep("r", 5)))
```

The Rdragmap run used 0.7% less aggregate CPU time but 10.2% more wall time.
Both runs exceeded physical-memory headroom and incurred millions of major
page faults, while CPU occupancy differed substantially. This is evidence of
CPU-work parity under a memory-constrained build, not evidence of a 10.2%
kernel regression; a larger-memory host and repeated counterbalanced runs are
required for a stable hash-build wall-time claim.

## Mapping compatibility and performance

A 100,000-pair prefix was mapped by both binaries against the Rdragmap-built
index using 16 threads and `--preserve-map-align-order true`. After removing
only the documented `@PG` and `@RG` header differences, all 200,146 SAM lines
were byte-identical with SHA-256
`50f9e6bce034f7d58e6952c8f16cc154f88e69caf23b85e82f67d1bc2a6a00af`.

```{r prefix-table, results='asis'}
prefix_rows <- subset(
  timings,
  phase == "mapping" & workload == "100000-pair prefix" & index_path == "compressed"
)
knitr::kable(
  format_timing_table(prefix_rows, "100,000-pair run"),
  align = c("l", rep("r", 5))
)
```

The 0.28-second wall difference is 0.47% in Rdragmap's favor and is below what
a single run can establish as a performance difference.

Both binaries then mapped all 10,110,535 pairs (20,221,070 reads). The complete
SAM streams were hashed without storing them; after removing only `@PG` and
`@RG`, both produced SHA-256
`2d84277058d45dd99861a39dbe55ad7c99be7477bce8c3a6cd9cbd3e5e735d60`.
The aligners reported 20,185,346 mapped reads (99.82%) and 20,190,926 total
alignments identically.

```{r full-table, results='asis'}
full_rows <- subset(
  timings,
  phase == "mapping" & workload == "complete exome" & index_path == "compressed"
)
knitr::kable(
  format_timing_table(full_rows, "Complete exome run"),
  align = c("l", rep("r", 5))
)
```

Rdragmap was 0.82 seconds (0.42%) faster in this single complete run and used
1.4% less aggregate CPU time. Treat this as measured parity rather than a
speedup claim. Hash decompression was also equivalent: 42.377 versus 42.650
seconds for the dominant auto-hit phase.

## Uncompressed memory-mapped index

A follow-up performance run on 2026-08-10 used the same Rdragmap executable,
index, inputs, thread count, and ordered-output setting. The upstream
`--ht-uncompress` operation materialized `hash_table.bin` and
`extend_table.bin` from the already validated compressed index, and alignment
used `--mmap-reference true`. The two additional files occupied about 28 GiB
and 3.2 GiB respectively; `hash_table.cmp` occupied 4.0 GiB.

```{r mmap-full-table, results='asis'}
mmap_full_rows <- subset(
  timings,
  phase == "mapping" & workload == "complete exome" &
    implementation != "untouched 4f98e00"
)
mmap_full_rows$implementation <- c(
  "compressed, per-run decompression",
  "uncompressed, memory-mapped"
)
knitr::kable(
  format_timing_table(mmap_full_rows, "Complete exome index path"),
  align = c("l", rep("r", 5))
)
```

The mmap run reported the same 20,221,070 input reads and 20,190,926 total
alignments. It sent SAM to `/dev/null`, so this is a performance and record-count
comparison rather than another full-output digest. On this host and workload,
mmap reduced wall time by 42.80 seconds (22.1%) and aggregate CPU time by
818.04 seconds (29.6%). It incurred 5,209 major page faults and reported a
134.51 thousand-read/s mapping rate, compared with 141.12 thousand reads/s for
the compressed run after its up-front decompression. The data therefore support
avoiding repeated decompression on this workload; they do not establish that
page-faulted mmap makes the mapping kernel itself faster.

The prefix results show the sensitivity of small-input timings to cache state
and the workflow's 100,000-pair scheduling unit:

```{r mmap-prefix-table, results='asis'}
mmap_prefix_rows <- subset(
  timings,
  phase == "mapping" & workload == "100000-pair prefix" &
    implementation != "untouched 4f98e00"
)
mmap_prefix_rows$implementation <- c(
  "compressed",
  "mmap, first run",
  "mmap, page-cache resident"
)
knitr::kable(
  format_timing_table(mmap_prefix_rows, "100,000-pair index path"),
  align = c("l", rep("r", 5))
)
```

### Ten-million-pair CPU and threading audit

The 100,000-pair profile was dominated by fixed startup and was not used for a
threading conclusion. On 2026-08-11, `perf record -F 99 -e cycles:u` instead
sampled both index paths over all **10,110,535 pairs** with 16 requested mapper
threads. A `/proc/PID/task/*/schedstat` sampler recorded each thread's executing
and run-queue time every 0.5 seconds. Linux reports this hybrid CPU's
efficiency-core and performance-core PMUs separately; percentages are within a
PMU and must not be added across PMUs.

```{r profile-table, results='asis'}
profile_rows <- subset(profiles, overhead_percent >= 3)
profile_table <- data.frame(
  `Index path` = profile_rows$index_path,
  PMU = profile_rows$pmu,
  Samples = format(
    profile_rows$sample_count,
    big.mark = ",",
    scientific = FALSE,
    trim = TRUE
  ),
  Symbol = profile_rows$symbol,
  Overhead = sprintf("%.2f%%", profile_rows$overhead_percent),
  check.names = FALSE
)
knitr::kable(profile_table, align = c("l", "l", "r", "l", "r"))
```

The complete compressed-path profile no longer resembles the startup-only
100,000-pair result. `autoHitThread` accounted for 27.05% and 33.03% of the two
PMUs, while Smith-Waterman, hash lookup, seeding, and alignment dominated the
remaining run. With mmap removing hash decompression, Smith-Waterman alone
accounted for 47.43% and 36.58% and hash-bucket processing for 9.21% and 11.25%.

The FASTQ gzip symbols were small even after mmap isolated the mapping phase:

```{r gzip-profile-table, results='asis'}
gzip_rows <- subset(
  profiles,
  index_path == "mmap" & symbol %in% c("inflate", "crc32_z")
)
gzip_table <- data.frame(
  PMU = gzip_rows$pmu,
  Symbol = gzip_rows$symbol,
  Overhead = sprintf("%.2f%%", gzip_rows$overhead_percent),
  check.names = FALSE
)
knitr::kable(gzip_table, align = c("l", "l", "r"))
```

Within each mmap PMU, `inflate` plus `crc32_z` totalled 1.79% and 4.25% of
sampled cycles. Even eliminating those cycles completely gives idealized
Amdahl bounds of only 1.8% and 4.4%; making them four times faster gives 1.4%
and 3.3%. `FastqNRecordReader::read` was 0.52%/0.89%, and `Tokenizer::reset` was
0.39%/0.42%. These are flat-profile cycle shares, not promises of wall
speedup, but they clearly reject gzip as the dominant CPU cost on this workload.

The scheduler samples show the surrounding concurrency rather than only where
executing cycles landed:

```{r threading-table, results='asis'}
threading_table <- data.frame(
  `Index path` = threading$index_path,
  Phase = threading$phase,
  `Peak live threads` = threading$peak_live_threads,
  `Elapsed sampled` = sprintf("%.1f s", threading$elapsed_seconds),
  `Mean cores executing` = sprintf("%.2f", threading$mean_core_equivalents),
  `Median cores executing` = sprintf("%.2f", threading$median_core_equivalents),
  `95th percentile` = sprintf("%.2f", threading$p95_core_equivalents),
  `Mean cores runnable` = sprintf("%.3f", threading$mean_runqueue_core_equivalents),
  check.names = FALSE
)
knitr::kable(
  threading_table,
  align = c("l", "l", rep("r", 6))
)
```

The source and runtime observations agree:

- compressed index loading uses `std::thread::hardware_concurrency()` rather
  than `--num-threads`; it reached 21 live threads including the main thread;
- mapping constructs a 32-thread pool so ordered blocks can park without
  preventing other work, but limits the outer executor to 16 mapper workers;
- the two gzip streams use independent but individually serial Boost/zlib
  filters, with a nested reader executor limited to two threads;
- input, insert-statistics feedback, alignment start, and output each retain an
  explicit block-order barrier at the inherited 100,000-pair scheduling unit;
- 16 threads accounted for more than 97% of sampled mmap cycles in each PMU;
  the other half of the pool was overwhelmingly parked;
- compressed mapping had a median 15.00 executing cores and a 16.00-core 95th
  percentile. The mmap profile's lower mean includes an initially cold mapped
  index with 157,417 major faults, plus pipeline ramp-up and drain, rather than
  evidence of FASTQ-reader saturation.

Therefore no rapidgzip or thread-broker integration is justified for this
pipeline now. Parallel gzip would have to take execution slots from the
alignment work that dominates the profile, while adding a substantial native
or Rust/FFI dependency and a new ordered-buffer lifecycle. The current
`rapidgzip` C++ design, `rapidgzip-rust` `Read + Send` decoder, and
`thread-broker` fixed-budget control law remain useful references if a future
production corpus shows material inflater share or sustained mapper starvation;
standalone decompressor throughput is not sufficient evidence to reopen that
decision.

The real run exposed one additional metrics-adapter defect unrelated to SAM
records: the internal `nextReference == -1` representation of same-contig SAM
`RNEXT =` was counted as a different chromosome. It falsely labelled all
20,165,314 mapped paired reads as interchromosomal. A focused regression now
covers the corrected sentinel interpretation. Re-running the 100,000-pair
prefix after the fix reported 444 of 199,412 mapped paired reads (0.22%) on
different chromosomes, including 215 (0.11%) at MAPQ 10 or greater.
